Insurance choice, decision assistance, and insurer steering

### 2021-05-03 finals/creed/creed-202105.Rmd
update slides for creed

- "More than half of all uninsured (56%) say they are likely to go to the marketplace to find a health plan in the future. But there are knowledge gaps."

- "Six in 10 are confused about or have not heard of the tax credit, and more than 7 in 10 want one-on-one help understanding their insurance options"

.footnote[
  [Robert Wood Johnson Foundation, **Understanding the Uninsured Now**, 2015](https://www.rwjf.org/en/library/research/2015/06/understanding-the-uninsured-now.html)
]

@@

1. [Background on the ACA](#background)
2. [Data](#data)
3. [Dominated choices](#dominated)
4. [Effects of decision assistance](#decision-assist)
5. [Insurer steering](#steering)

---
class: inverse, center, middle
name: background

# Background on the ACA exchanges

---
class: clear, center, middle

![:scale 80%](https://media.giphy.com/media/Awb1k8lX8a3Re/giphy.gif)

---
# What are the exchanges?

- Online marketplace to purchase private health insurance for those without insurance through their employers or other public programs

- Intended to help fill the "coverage gap" for the 40+ million uninsured individuals in the U.S. (pre ACA)

- Complicated program with various premium and cost-sharing subsidies, specific to a household

---
# Metal tiers

.pull-left[
To help with this complex decision, plans organized around tiers based on actuarial value
]

.pull.right[
  ![:scale 50%](../pics/metal_tiers.jpg)
]

---
# Community rating rules

.pull-left[
Insurers limited in ability to adjust premiums

- Smoking (not in CA)
- Age
- Rating area
]

.pull-right[
  ![:scale 100%](../pics/CA_Rating_Regions2.jpg)
]

---
# Popularity of the exchanges

.center[
![:scale 85%](../pics/aca-enrollment-graph.png)
]

@@

Main enrollment come from Covered California under FOIA. This includes:
- Individual demographics and income
- Plan purchases
- Information on type of assistance (if any)

@@

---
# Final data

All analysis done at the household level...

- **Initial data:** 16.9 million households from 2014-2019
- Remove households with missing or inconsistent data
  - 14.6 million remaining
- Remove households predicted to be insured elsewhere
  - 8.3 million remaining
  - 4.5 million of which are new enrollees

---
# Enrollees over time

.center[
  ![:scale 50%](../../figures/enrollee_count.png)
]

@@

- Denote by $Pr[Y_{ij}^{1}]$ the probability of household $i$ selecting plan $j$ with decision assistance

- Denote by $Pr[Y_{ij}^{0}]$ the probability of household $i$ selecting plan $j$ without decision assistance

@@

$$ATT = Pr[Y_{ij}^{1} | \text{Assist}] - Pr[Y_{ij}^{0} | \text{Assist}]$$

@@

2. Use $\hat{\theta}$ to form predicted counterfactual, $\widehat{Pr[Y_{ij}^{0} | \text{Assist}]}$

3. Use predicted value in ATT estimate: $$\underbrace{Pr[Y_{ij}^{1} | \text{Assist}]}_{Observed} - \underbrace{\widehat{Pr[Y_{ij}^{0} | \text{Assist}]}}_{Predicted}$$

4. Aggregate to level of insurer and metal

@@

>>>>

@@

1. [Data](#data)
2. [Dominated choices](#dominated)
3. [Effects of decision assistance](#decision-assist)
4. [Insurer steering](#steering)

@@

Main enrollment come from Covered California under FOIA. This includes:
- Individual demographics and income
- Plan purchases
- Information on type of assistance (if any)
- 8.3 million households, 4.5 million new enrollees

@@

@@

- Denote by $N_{j}^{1} = \sum_{i} 1(Y_{ij}^{1}=1 | \text{Assist})$ the the count of enrollees in plan $j$ had they received decision assistance, conditional on receiving assistance

- Denote by $N_{j}^{0} = \sum_{i} 1(Y_{ij}^{0}=1 | \text{Assist})$ the count of enrollees in plan $j$ had they not received decision assistance, conditional on receiving assistance

@@

$$ATT = N_{j}^{1} - N_{j}^{0} $$

@@

2. Use $\hat{\theta}$ to form predicted counterfactual, $\hat{N}_{j}^{0} = \sum_{i} \widehat{Pr[Y_{ij}^{0} | \text{Assist}]}$

3. Use predicted value in ATT estimate: $$\underbrace{N_{j}^{1}}_{Observed} - \underbrace{\hat{N}_{j}^{0}}_{Predicted}$$

4. Aggregate to level of insurer and metal tier

@@

---
# Selection on observables

.pull-left[
  ![](../../figures/cov_balance.png)
]

.pull-right[
  ![](../../figures/ps_assist_full.png)
